---
title: "Comparing species composition survey methods. Part 2a: Species Lists and Abundance Curves"
author: "Kelly Carscadden"
date: '2020-03-04'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 2. Community composition

#### Species Lists and Abundance Curves
```{r, include=F}
library(tidyverse)
library(forcats)

splong <- read_csv("splong_NWT_bs.csv")[-1]
```

```{r, include=F}
#species counts for all hits
tallyf <- splong %>% count(sp) %>% arrange(desc(n))
head(tallyf)

#species counts for top hits only
h1 <- splong %>% filter(hits == "hit_1")
tally1 <- h1 %>% count(sp)

# note, diffs in sp number overall = sp lists not comparable
```

```{r, echo=F}
# total species from all hits
n_all <- length(unique(tallyf$sp))

# total species from top hits only
n_top <- length(unique(tally1$sp))
```

Total species from *all hits*: `r n_all`

Total species from *top hits*: `r n_top`

Which species were totally missed by the top hits survey approach?

```{r, echo=F}
# missing sp
(miss <- tallyf$sp[which(!tallyf$sp %in% tally1$sp)])
```

How do abundance curves compare across the two survey methods (across sites, treatments)?
```{r, echo=F}
# add 0s for those to make sp lists comparable
adds <- data.frame(sp=miss, n=0)
tally1 <- rbind(tally1, adds)

# order species by decreasing abundance
tallyf <- tallyf %>% mutate(sp = as.factor(sp),
                            sp = fct_reorder(sp, -n)) 

# match order of sp to the all-hits method above
tally1 <- tally1 %>% mutate(sp = as.factor(sp))
tally1 <- tally1[order(match(tally1$sp, tallyf$sp)),]

# stacking data for easier plotting
tallyf$method <- "All hits"
tally1$method <- "Top hits"
tally_combo <- rbind(tallyf, tally1)

# plot  
tally_combo %>% ggplot(aes(x=sp, y=n, col=method))+
geom_hline(yintercept = 0, lty=2, col="lightgrey")+
  geom_point(alpha=0.6) +
  theme_classic()+
  theme(axis.text.x = element_text(angle=90, size=5))+
  labs(x="Species", y="Total abundance")+
  scale_color_manual(values = c("black", "cornflowerblue"), name="Survey method")
```

Across methods, abundances seem quite consistent. This plot suggests the main impact of top hits is:

(a) decreasing the frequency of records of the most abundant species, and
(b) losing a few of the rarest species


Does a top hit approach miss a consistent proportion of each species' occurrences, or disproportionately miss certain species?

Formula: (tally_all_hits - tally_top_hits) / tally_all_hits

```{r, echo=F}
# since the sp order is consistent across these df, can subtract abundances
tallyf$loss<- tallyf$n - tally1$n
tallyf$prop_loss <- tallyf$loss/tallyf$n

# plot
tallyf %>% ggplot(aes(x=sp, y=prop_loss))+
  geom_hline(yintercept = 0, lty=2, col="lightgrey")+
  geom_hline(yintercept = 1, lty=2, col="lightgrey")+
  geom_point(alpha=0.6) +
  theme_classic()+
  theme(axis.text.x = element_text(angle=90, size=5))+
  labs(x="Species", y="Proportion missed")
```

Species are displayed in order of decreasing abundance, as in the previous plot. As expected, rarer species vary in how they're impacted (e.g., since one missed observation can mean 100% loss of that species, or conversely one top hit of a taller rare sp can be all that was seen, and lower canopy surveys would not have added more observations anyways).

Proportion loss varies across species (i.e., is not that you consistently lose x% of all observations across the board, by using the top hits approach). So, the effects of top hits could be challenging to account for post-hoc.
