---
title: "Cleaning NWT Black Sand data"
author: "Kelly Carscadden"
date: '2019-12-10'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Black Sand data munging 
- fixing subplot names
- fixing sp names
- converting to long data format

2018 Black Sand data downloaded from Laurel Brigham email, Nov 25, 2019

```{r}
library(tidyverse)

# load data
spbs <- read_csv("Subplots_QAQC.csv")

# peek at data
str(spbs)
head(spbs)
```

### 1. Standardizing subplot naming
(eg., A CO -> ACO, Audubon Control -> ACO)
```{r}
# look at unique subplot labels
unique(spbs$subplot)

# format: replace(subplot, existing_name, replacement)
spbs <- spbs %>% 
  mutate(subplot = str_replace(subplot, "A CO", "ACO"),
        subplot = str_replace(subplot, "Audubon BS", "ABS"),
        subplot = str_replace(subplot, 
                              "Audubon Control", "ACO"),
        subplot = str_replace(subplot, 
                              "East Knoll Control", "ECO"),
        subplot = str_replace(subplot, "Soddie BS", "SBS"),
        subplot = str_replace(subplot, 
                              "Soddie Control", "SCO"),
        subplot = str_replace(subplot, "T CO", "TCO")
) #end mutate

unique(spbs$subplot) #good
```

### 2. Standardizing species naming

- Remove sp that weren't hit (listed as P under points)
```{r}
# creating black sand species list
spbs %>% 
  #remove sp that weren't hit (but are in plot)
  filter(point != "P") %>%
  #remove extraneous columns
  select(subplot:hit_8) %>%
  #melt so have hits as 1 col, sp as another
  pivot_longer(-c(subplot,point), names_to = "hits", 
               values_to = "sp") %>%
  distinct(sp) %>%
  #alphebetize so easier to see any other inconsistencies
  arrange(sp) %>%
  print(n=1000) #print all rows
```

- Remove any IDs that aren't vascular plants (e.g., rock, scat)

- Fix inconsistencies in sp names, with reference to NWT species codes (pspecies.mw.data_JGS.csv):
    - Uncertain what LI is, removed since not consistent with sp names
    - Some sp IDs uncertain...Removed "?" and "*" and " U" at name end to synonymize with confident IDs of the same sp

```{r}
# filtering sp IDs

# list the terms to remove
elim <- c("MOSS", "Moss", "LITTER", "SCAT", "Scat", 
          "Elk scat", "ROCK", "BARE", "LI", "LICHEN")

splong <- spbs %>% filter(point != "P") %>%
  select(subplot:hit_8) %>%
  pivot_longer(-c(subplot,point), names_to = "hits", 
               values_to = "sp") %>%
  #drop names to eliminate
  filter(! sp %in% elim) %>%
  #drop NA
  drop_na() %>%
  #remove ?, *, and U terms at end of some sp names
  #fix sp IDs
  mutate(sp = str_remove(sp, pattern="[?*]"),
         sp = str_remove(sp, pattern=" U"),
         sp = str_replace(sp, "CAR sp.", "CARSPP"),
         sp = str_replace(sp, "CAR sp.", "CARSPP"),
         sp = str_replace(sp, "CAR sp. 1", "CARSPP1"),
         sp = str_replace(sp, "CARSPP 1", "CARSPP1"),
         sp = str_replace(sp, "CAR sp. 2", "CARSPP2"),
         sp = str_replace(sp, "Car sp. 2", "CARSPP2"),
         sp = str_replace(sp, "Campanula", "CAMPSP"),
         sp = str_replace(sp, "CAM sp.", "CAMSPP"),
         sp = str_replace(sp, "CAMPSP", "CAMSPP"),
         sp = str_replace(sp, "CARDES", "CARROS"),
         sp = str_replace(sp, "CARDAS", "CARROS"),
         sp = str_replace(sp, "CASSOC", "CASOCC"),
         sp = str_replace(sp, "CHAM sp.", "CHASPP"),
         sp = str_replace(sp, "CALPER", "CALPUR"),
         sp = str_replace(sp, "COBMYO", "KOBMYO"),
         sp = str_replace(sp, "DESCAE", "DESCES"),
         sp = str_replace(sp, "ERIMIL", "ERIMEL"),
         sp = str_replace(sp, "ERININ", "ERINAN"),
         sp = str_replace(sp, "ERISUM", "ERISIM"),
         sp = str_replace(sp, "ERISUM U", "ERISIM U"),
         sp = str_replace(sp, "GUNDRU", "JUNDRU"),
         sp = str_replace(sp, "LEWPIG", "LEWPYG"),
         sp = str_replace(sp, "LL sp.", "LLOSER"),
         sp = str_replace(sp, "LLSPP", "LLOSER"),
         sp = str_replace(sp, "LLUSPI", "LUZSPI"),
         sp = str_replace(sp, "MINOPT", "MINOBT"),
         sp = str_replace(sp, "PHALP", "PHLALP"),
         sp = str_replace(sp, "PHALP?", "PHLALP"),
         sp = str_replace(sp, "PHLPOL", "PHLPUL"),
         sp = str_replace(sp, "POA sp.", "POASPP"),
         sp = str_replace(sp, "POA fat", "POAFAT"),
         sp = str_replace(sp, "SEDLEN", "SEDLAN"),
         sp = str_replace(sp, "SELDAN", "SELDEN"),
         sp = str_replace(sp, "SOLMOL", "SOLMUL"),
         sp = str_replace(sp, "VAC sp.", "VACSPP")
         ) #end mutate
```

Check revised species list
```{r}
splong %>%
  distinct(sp) %>%
  #alphebetize
  arrange(sp) %>%
  print(n=1000)

#write.csv(splong, "splong_NWT_bs.csv")
```

